---
title: "Assignment 3: Data Exploration"
author: "Fiona Price"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
#Load the dplyr, tidyverse, ggplot2, and here package
library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)

neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE)  
litter <- read.csv(
  file = here("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)  

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.
 2.
 3. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(neonics)
#The dataset has 4623 rows and 30 columns.
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
#Use the summary function to look at an overview of the types of effects.
summary(neonics$Effect)

#Create a table summarizing these effects.
neonics_summary <- table(neonics$Effect)

#Now, sort the table using the sort function. Decreasing = TRUE indicates that
#the sort should be decreasing (so it starts with the most common effect).
sorted_neonics_summary <- sort(neonics_summary, decreasing = TRUE)
sorted_neonics_summary

#Population is the most common effect. Mortality is the second most common.
```

> Answer:


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
#Use the summary function to look at an overview of the common species names.
summary(neonics$Species.Common.Name)

#Create a table summarizing these species.
neonics_species <- table(neonics$Species.Common.Name)

#Now, sort the table using the sort function. Decreasing = TRUE indicates that
#the sort should be decreasing (so it starts with the most common species).
sorted_neonics_species <- sort(neonics_species, decreasing = TRUE)
sorted_neonics_species

#The six most commonly studied species are the honey bee, parasitic wasp, buff
#tailed bumblebee, carniolan honey bee, bumble bee, and Italian honeybee. 

#This is another way to do it using maxsum. Here, I'm indicating I want seven
#levels to show (because the 7th will be (other)).
summary(neonics$Species.Common.Name, maxsum = 7)
```

> Answer:

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}

view(neonics$Conc.1..Author.)
class(neonics$Conc.1..Author.)
#The class is factor.

```

> Answer:

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(neonics, aes(Publication.Year)) +
         geom_freqpoly()

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(neonics, aes(Publication.Year, color = Test.Location)) +
         geom_freqpoly() +
  labs(color = "Test Location") 

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(neonics, aes(Endpoint)) + #Our data is neonics; our x-axis is endpoint.
  geom_bar() + #This indicates we want a bar plot.
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #This 
#rotates and aligns and the x-axis labels so we can see them clearly. 

```

> Answer: The two most common endpoints are LOEL and NOEL. LOEL is defined as
the Lowest-observable-effect-level: lowest dose (concentration) producing 
effects that were significantly different (as reported by authors) from 
responses of controls (LOEAL/LOEC). NOEL is defined as 
no-observable-effect-level: highest dose (concentration) producing effects not 
significantly different from responses of controls according to author's 
reported statistical test (NOEAL/NOEC).

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(litter$collectDate)
litter$collectDate <- as.Date(litter$collectDate)
class(litter$collectDate)
unique(litter$collectDate)

```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
#Use unique to determine how many different plots were sampled at Niwot Ridge.
unique(litter$namedLocation)
#There are 12 levels, so 12 different plots were sampled.

#Use summary to determine how many different plots were sampled at Niwot Ridge.
summary(litter$namedLocation)
#This returned the sample counts at each plot.
```

> Answer: `unique` returns the number of levels of our data (12). `summary` 
returns the number of counts of each level. We can still find the number of 
levels using summary by manually counting the number of categories in the 
output.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
